Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-26 VMCAI-revised-version/final.tex
up

This all changes when one considers VASS (having control-states). For a VASS $V = (Q,T)$, a set $\Bad = \mathop{\uparrow} \Bad$, for all $q \in Q$, either there is an element of the basis of $\Bad$ with control-state $q$ \----
then $q(\textbf{v})$ with $\textbf{v} > \textbf{v}_{\Bad}$ is necessarily in $\Bad$ by upward closure; if this hold for all $q \in Q$, then $k$-resilience does not hold \---- either there is none. If there is none, then 
$\mathop{\uparrow} q(\textbf{0})$
 is in the complement of $\Bad$, i.e. $\Safe$.
Based on this, we partition $\Safe$ into two subsets: an upward-closed 
union of sets of the form $\mathop{\uparrow} q(\textbf{0})$, and a remaining downward-closed subset.
The key now is that
there is only a finite number of elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} \leq \textbf{u}_k$ for all of which we will check reachability of $\Safe$ in $k$ or less steps, while 
 elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} > \textbf{u}_k$ can only potentially reach the upward-closed subset of $\Safe$ in $k$ or less steps. 
Only being concerned by the upward-closed subset of $\Safe$ in the latter case enable us to reuse techniques seen when $\Safe$ is upward-closed. Indeed we can compute the basis for the sets of elements from which our upward-closed subset of $\Safe$ is reachable in at most $k$ steps. 
Substracting these predecessor from $\Bad$ yields either a finite number of elements from which one has to check $\Safe$ is reachable in at most $k$ steps, either an infinite number of elements of which there is one which cannot reach $\Safe$ in at most $k$ steps \-- for much the same reasons {\sc Bounded resilience} and {\sc $k$-resilience} never hold for VAS when $\Safe = \mathop{\downarrow} \Safe$ and $\Bad = \mathop{\uparrow} \Bad$. This leads to a decision procedure detailed in appendix~\ref{appendix vass},
 which lead to the following decidability result:
\begin{theorem}\label{vass down}
{\sc $k$-resilience }  and {\sc Bounded resilience} are decidable for VASS when 
$\Safe = \mathop{\downarrow} \Safe$.
\end{theorem}

@@

From $k_0$ onwards,
$\Bad \setminus (\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))= 
\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is stationnary. 
%\mathieu{
We check the finiteness of the set by comparing the elements from the finite basis of $\Bad$ and
$(\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$.
%}
If the set is finite then we stop the procedure once we have checked for every element
whether it is possible to reach $\Safe$ from it.
If it is possible to reach $\Safe$ from every element then {\sc $k_m$-resilience}
hold
for $k_m = \max(k_0, k_{\pi})$ with 
$k_{\pi}$ the maximum of the length of the paths from 
$\Bad \setminus (\pred^{\leq k_0}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k_0}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$ to $\Safe$.
If it is infinite then
for every $k \in \N$,
$k$-resilience does not hold, 
and we eventually find a witness that $k$-resilience does not hold.
Indeed, 
if
$\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \cup \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is
infinite
then
there exists an element of the form $q(\textbf{u} )$
with $\textbf{u} > \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, \ldots, c_{\max})$ 
that is in $\Bad$ but not in any of the $\pred^{\leq k}(\mathop{\uparrow} q_{j_i}(\textbf{0}))$,
and hence from which it is not possible at all to reach
$\Safe$ in $k$ or less steps.
\qed
\end{proof}

>>>>

This all changes when one considers VASS (having control-states). For a VASS $V = (Q,T)$, a set $\Bad = \mathop{\uparrow} \Bad$, for all $q \in Q$, either there is an element of the basis of $\Bad$ with control-state $q$ \----
then $q(\textbf{v})$ with $\textbf{v} > \textbf{v}_{\Bad}$ is necessarily in $\Bad$ by upward closure; if this hold for all $q \in Q$, then $k$-resilience does not hold \---- either there is none. If there is none, then 
$\mathop{\uparrow} q(\textbf{0})$
 is in the complement of $\Bad$, i.e. $\Safe$.
Based on this, we partition $\Safe$ into two subsets: an upward-closed 
union of sets of the form $\mathop{\uparrow} q(\textbf{0})$, and a remaining downward-closed subset.
The key now is that
there is only a finite number of elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} \leq \textbf{u}_k$ for all of which we will check reachability of $\Safe$ in $k$ or less steps, while 
 elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} > \textbf{u}_k$ can only potentially reach the upward-closed subset of $\Safe$ in $k$ or less steps. 
Only being concerned by the upward-closed subset of $\Safe$ in the latter case enable us to reuse techniques seen when $\Safe$ is upward-closed. Indeed we can compute the basis for the sets of elements from which our upward-closed subset of $\Safe$ is reachable in at most $k$ steps. 
Substracting these predecessor from $\Bad$ yields either a finite number of elements from which one has to check $\Safe$ is reachable in at most $k$ steps, either an infinite number of elements of which there is one which cannot reach $\Safe$ in at most $k$ steps \-- for much the same reasons {\sc Bounded resilience} and {\sc $k$-resilience} never hold for VAS when $\Safe = \mathop{\downarrow} \Safe$ and $\Bad = \mathop{\uparrow} \Bad$.
We check the finiteness of the set by comparing the elements from the finite basis of both upward-closed sets. This leads to a decision procedure detailed in appendix~\ref{appendix vass},
 which lead to the following decidability result:
\begin{theorem}\label{vass down}
{\sc $k$-resilience }  and {\sc Bounded resilience} are decidable for VASS when 
$\Safe = \mathop{\downarrow} \Safe$.
\end{theorem}

@@

From $k_0$ onwards,
$\Bad \setminus (\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))= 
\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is stationnary. 
%\mathieu{
We check the finiteness of the set by comparing the elements from the finite basis of $\Bad$ and
$(\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$.
%
As an example, in dimension $2$, for an upward-closed set $U$, $\mathop{\uparrow} q(n,m) \setminus U$ is finite if and only if $U$ contains an element of the form $q(l,m')$ with $m' \geq m$ and $l \leq n$
and one of the form $q(n',l)$ with $n' \geq n$ and $l \leq m$. Indeed, if it does, then
elements of $\mathop{\uparrow} q(n,m)$ with coordinates sufficiently large are greater than these and thus belong to $U$.
If not, then the infinite set $\{q(n,m') \mid m' \geq m \}$ or the infinite set
$\{q(n',m) \mid n' \geq n \}$   belong to $\mathop{\uparrow} q(n,m) \setminus U$ which is hence not finite.
%}
If the set is finite then we stop the procedure once we have checked for every element
whether it is possible to reach $\Safe$ from it.
If it is possible to reach $\Safe$ from every element then {\sc $k_m$-resilience}
hold
for $k_m = \max(k_0, k_{\pi})$ with 
$k_{\pi}$ the maximum of the length of the paths from 
$\Bad \setminus (\pred^{\leq k_0}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k_0}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$ to $\Safe$.
If it is infinite then
for every $k \in \N$,
$k$-resilience does not hold, 
and we eventually find a witness that $k$-resilience does not hold.
Indeed, 
if
$\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \cup \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is
infinite
then
there exists an element of the form $q(\textbf{u} )$
with $\textbf{u} > \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, \ldots, c_{\max})$ 
that is in $\Bad$ but not in any of the $\pred^{\leq k}(\mathop{\uparrow} q_{j_i}(\textbf{0}))$,
and hence from which it is not possible at all to reach
$\Safe$ in $k$ or less steps.
\qed
\end{proof}
